---
title: "Preparing and Investigating Data"
teaching: 60
exercises: 0
questions:
- "What to do before you can bring the data into R?"
objectives:
- "Identify different data types"
- "Recognise different types of visualisations"
keypoints:
- "Data should have 1 observation per row."
source: "Rmd"
---

```{r setup, include=FALSE}
source("../bin/chunk-options.R")
```

First, we need to load the `tidyverse` package (actually a collection of packages).

```{r packages, message=FALSE}
library(tidyverse)
```

We can now load the **Pattani** data from the Excel file. For the notes, this is stored
in the `data` directory. It might be different for you.

```{r read_data}
path <- file.path("..", "data", "Pattani.xlsx")
pattani <- readxl::read_excel(path, sheet = "Data", skip = 1)
```

`readxl::read_excel` is used to explicitly call the `read_excel` function in the `readxl`
package.

We would like to see what the data looks like now that it has been read into R. You can
just type the name of the dataset.

```{r print_pattani}
pattani
```

> ## Read in the cuckoo data
>
> Use the `read_csv` function in the `readr` package to load `cuckoo.csv`.
>
> > ## Solution
> >
> > ```{r read_cuckoo, message=FALSE}
> > cuckoo <- readr::read_csv(file.path("..", "data", "cuckoo.csv"))
> > cuckoo
> > ```
> {: .solution}
{: .challenge}

Having names with spaces in them can make them more difficult to work with. We can add to
the way that we read the file to modify the names.

```{r rename_columns}
pattani <- readxl::read_excel(path, sheet = "Data", skip = 1) %>%
  rename(blood_lead = `blood lead`, ln_blood_lead = `ln(blood lead)`)

pattani
```

> ## What does `%>%` mean?
> 
> `%>%` is something from the `tidyverse` called a **pipe**. It allows you to take the output
> from 1 function (in this case `read_excel`) and put it directly into another function
> (`rename`). It is as though there is a **pipe** connecting the functions together.
{: .callout}

### Simple summary statistics

Previously we got a preview of the data, but we might want to get some summary statistics
about the columns. You can use the function `summary` for this.

```{r summary}
summary(pattani)
```

Have a look at the `character` columns. What sort of variables are these?

If we convert these to `factor` then R knows that they are categorical variables and does
some helpful things for us. We will use the `as_factor` function, which is part of the
`tidyverse`. There is also a base R function `as.factor` (note the `.` instead of `_`)
but we will use the `tidyverse` version.

```{r convert_to_factors}
pattani <- readxl::read_excel(path, sheet = "Data", skip = 1) %>%
  rename(blood_lead = `blood lead`, ln_blood_lead = `ln(blood lead)`) %>%
  mutate(gender = as_factor(gender), school = as_factor(school), water = as_factor(water))
```

What happens if we run the summary now?

```{r summary_factors}
summary(pattani)
```

The factor variables now show the number of each element in the variables.

> ## Convert cuckoo data to a factor
>
> Convert the `Nest` column to a factor and display the summary.
>
> > ## Solution
> >
> > ```{r cuckoo_factor, message=FALSE}
> > cuckoo <- readr::read_csv(file.path("..", "data", "cuckoo.csv")) %>%
> >   mutate(Nest = as_factor(Nest))
> > summary(cuckoo)
> > ```
> {: .solution}
{: .challenge}

## Visualising the data

We will start with a histogram of the levels of lead in the blood.

```{r histogram_bl, warning=FALSE, message=FALSE}
ggplot(pattani, aes(x = blood_lead)) +
  geom_histogram()
```

I am not a fan of the grey background that is the default with `ggplot` graphs. You can
change that by adding a theme.

```{r histogram_bl_bw, warning=FALSE, message=FALSE}
ggplot(pattani, aes(x = blood_lead)) +
  geom_histogram() +
  theme_bw()
```

I will set the default theme so that I don't have to specify it each time.

```{r set_theme}
theme_set(theme_bw())
```

> ## Create a histogram of the cuckoo egg lengths
>
> Create a histogram of the cuckoo egg lengths
>
> > ## Solution
> >
> > ```{r cuckoo_histogram}
> > ggplot(cuckoo, aes(x = Length)) +
> >   geom_histogram(bins = 10)
> > ```
> {: .solution}
{: .challenge}

### Distribution?

> - The **distribution** of a variable tells us what values it takes and how often it
>   takes these values.
> - When examining a distribution, look for **shape**, **center**, and **spread** and for
>   clear **deviations** from the overall shape.
>
> _Moore, McCabe & Craig (2017)_
{: .quotation}

You have probably heard of the Normal distribution, which has a **bell shape**.

```{r normal_dist, fig.cap="Normal distribution"}
ggplot(data = data.frame(x = c(-3, 3)), aes(x)) +
  stat_function(fun = dnorm, n = 101, args = list(mean = 0, sd = 1)) + ylab("") +
  scale_y_continuous(breaks = NULL)
```

There are several other distributions. Several of the tests and models that we will discuss
in this workshop have as one of the assumptions that either the data or the residuals
of the model are Normally distributed.

### Now back to what we were doing

For a lot of the statistical tests that we will be doing later, we would like the variable
to be normally distributed. Looking at the histogram, it should be symmetric and bell-shaped.

```{r histogram_lnbl, warning=FALSE, message=FALSE}
ggplot(pattani, aes(x = ln_blood_lead)) +
  geom_histogram()
```

It looks a little bit better. We can also use a QQ-plot to check for normality. The points
should lie approximately along the line.

```{r qqplot, warning=FALSE}
ggplot(pattani, aes(sample = blood_lead)) +
  geom_qq() +
  geom_qq_line()
```

```{r qqplot_ln, warning=FALSE}
ggplot(pattani, aes(sample = ln_blood_lead)) +
  geom_qq() +
  geom_qq_line()
```

What if we wanted to see the difference between the boys and girls?

```{r histogram_gender, warning=FALSE, message=FALSE}
ggplot(pattani, aes(x = blood_lead, fill = gender)) +
  geom_histogram()
```

That produced a stacked histogram, which isn't all that easy to read. It might be better
if they were side-by-side.

```{r histogram_gender_dodge, warning=FALSE, message=FALSE}
ggplot(pattani, aes(x = blood_lead, fill = gender)) +
  geom_histogram(position = "dodge")
```

> ## Cuckoo egg lengths by host nest
>
> Create a histogram of the cuckoo egg lengths with colours showing the host nest species.
>
> > ## Solution
> >
> > ```{r cuckoo_histogram_nest}
> > ggplot(cuckoo, aes(x = Length, fill = Nest)) +
> >   geom_histogram(position = "dodge", bins = 10)
> > ```
> {: .solution}
{: .challenge}

There doesn't seem to be much difference. Perhaps we should try a box plot. Notice that
`gender` is now the `x` variable and `blood_lead` is the `y` variable.

```{r boxplot_gender, warning=FALSE}
ggplot(pattani, aes(x = gender, y = blood_lead)) +
  geom_boxplot()
```

The box plot also suggests that there isn't much difference between boys and girls.
What about between schools?

```{r boxplot_school, warning=FALSE}
ggplot(pattani, aes(x = school, y = blood_lead)) +
  geom_boxplot()
```

> ## Box plot of cuckoo egg lengths by host nest
>
> Create a box plot of the cuckoo egg lengths with colours showing the host nest species.
>
> > ## Solution
> >
> > ```{r cuckoo_boxplot}
> > ggplot(cuckoo, aes(x = Nest, y = Length)) +
> >   geom_boxplot()
> > ```
> {: .solution}
{: .challenge}

There does seems to be more difference between schools than between genders. We can add
gender back in again. Again, we have to use `position = "dodge"`.

```{r boxplot_school_gender, warning=FALSE}
ggplot(pattani, aes(x = school, y = blood_lead, fill = gender)) +
  geom_boxplot(position = "dodge")
```

You can add a little bit of space between the boxes.

```{r boxplot_school_gender_dodge, warning=FALSE}
ggplot(pattani, aes(x = school, y = blood_lead, fill = gender)) +
  geom_boxplot(position = position_dodge(1))
```


{% include links.md %}
